---
title: "01 — Basak snRNA-seq: Loading, QC, Integration, and Annotation"
---

This script processes the four Basak snRNA-seq Chromium runs (LC001A, LC001B,
LC002A, LC002B). It produces the annotated, QC-passed Seurat object used in all
downstream analyses.

Output: `output/01_basak/basak_snRNA_seurat.rds` (2,071 nuclei)

```{r setup}
library(Seurat)
library(dplyr)
library(ggplot2)

source("R/qc_plots_by_sample.R")
source("R/apply_qc_thresholds.R")

dir.create("output/01_basak", recursive = TRUE, showWarnings = FALSE)
```


# 1. Load raw count matrices

```{r load}
LC001A <- CreateSeuratObject(
  counts = Read10X(data.dir = "data/snRNA/LC001A_filtered_feature_bc_matrix"),
  project = "LC"
)
LC001A[["percent.mt"]] <- PercentageFeatureSet(LC001A, pattern = "^MT-")
LC001A$Section      <- "LC001A"
LC001A$SectionLevel <- "Rostral"

LC001B <- CreateSeuratObject(
  counts = Read10X(data.dir = "data/snRNA/LC001B_filtered_feature_bc_matrix"),
  project = "LC"
)
LC001B[["percent.mt"]] <- PercentageFeatureSet(LC001B, pattern = "^MT-")
LC001B$Section      <- "LC001B"
LC001B$SectionLevel <- "Rostral"

LC002A <- CreateSeuratObject(
  counts = Read10X(data.dir = "data/snRNA/LC002A_filtered_feature_bc_matrix"),
  project = "LC"
)
LC002A[["percent.mt"]] <- PercentageFeatureSet(LC002A, pattern = "^MT-")
LC002A$Section      <- "LC002A"
LC002A$SectionLevel <- "Caudal"

LC002B <- CreateSeuratObject(
  counts = Read10X(data.dir = "data/snRNA/LC002B_filtered_feature_bc_matrix"),
  project = "LC"
)
LC002B[["percent.mt"]] <- PercentageFeatureSet(LC002B, pattern = "^MT-")
LC002B$Section      <- "LC002B"
LC002B$SectionLevel <- "Caudal"

# Merge all four runs
basak <- merge(LC001A, y = c(LC001B, LC002A, LC002B))
table(basak$Section, useNA = "ifany")   # LC001A=497 LC001B=540 LC002A=629 LC002B=681
rm(LC001A, LC001B, LC002A, LC002B)
```


# 2. Pre-QC integration (all cells, including low-quality)

Integration before QC is required to define QC batches per cell type and sample.

```{r preqc-integrate}
basak$log2_nCount_RNA <- log2(basak$nCount_RNA + 1)

basak <- NormalizeData(basak, verbose = FALSE)
basak <- FindVariableFeatures(basak, verbose = FALSE)
basak <- ScaleData(basak, verbose = FALSE)
basak <- RunPCA(basak, verbose = FALSE)
basak <- IntegrateLayers(basak, method = RPCAIntegration,
                         verbose = FALSE, k.weight = 100)
basak <- RunUMAP(basak, dims = 1:40, reduction = "integrated.dr",
                 verbose = FALSE)
basak <- FindNeighbors(basak, reduction = "integrated.dr",
                       dims = 1:40, k.param = 20, verbose = FALSE)
basak <- FindClusters(basak, resolution = seq(0.1, 1.5, by = 0.1),
                      verbose = FALSE)
basak[["RNA"]] <- JoinLayers(basak[["RNA"]])
```


# 3. Preliminary broad annotation (to define QC batches)

Cell types are annotated at res 0.5 based on canonical marker expression.
NE neurons: DBH, TH, SLC6A2; oligodendrocytes: MAG, MBP, MOG, PLP1;
OPCs: PDGFRA, EGFR; astroependymal: GFAP, GJA1, AQP4, FOXJ1;
microglia: C1QB, AIF1. Cluster 6 is a low-quality cluster (low UMI, high MT%).

```{r preqc-annotate}
Idents(basak) <- basak$RNA_snn_res.0.5

basak <- RenameIdents(basak,
  "0"  = "Neurons_NE",
  "1"  = "Neurons",
  "3"  = "Neurons",
  "7"  = "Neurons",
  "10" = "Neurons",
  "2"  = "Oligo",
  "4"  = "Oligo",
  "9"  = "OPCs",
  "5"  = "Astroependymal",
  "8"  = "Microglia",
  "6"  = "low_UMI"
)

basak$CellType_AllCells <- Idents(basak)
```


# 4. QC: data-driven, batch-aware thresholds

QC batches were defined as Section × cell type for larger populations 
(NE neurons, other neurons, oligodendrocytes), while small cell-type populations 
(OPCs, astroependymal, microglia) and the low-UMI cluster were pooled across 
sections into a single batch each.

The `qc_plots_by_sample` function estimates per-batch thresholds from kernel 
density estimates of log2(nCount_RNA) and log2(nFeature_RNA), aided by Otsu 
thresholds and Gaussian mixture modelling where applicable. Per-batch 
thresholds are reported in Supplementary Table 1.

```{r qc-run}
basak$qc_batch <- paste(basak$Section, basak$CellType_AllCells, sep = "_")
basak$qc_batch <- ifelse(
  basak$CellType_AllCells %in% c("OPCs", "Astroependymal", "Microglia", "low_UMI"), 
  as.character(basak$CellType_AllCells), 
  basak$qc_batch
  )

res <- qc_plots_by_sample(
  seurat_obj     = basak,
  sample_col     = "qc_batch",
  log2_qc_cols   = c("nFeature_RNA", "nCount_RNA"),
  qc_cols        = c("percent.mt"),
  path           = "output/01_basak/QC_plots/",
  group.by       = "Section",
  upper_quantile = 0.997,
  mt_max         = 100,
  valley_rules   = list(.default = list(strategy = "auto")),
  floors_raw     = list(nFeature_RNA = 300),
  plots_ncol     = 4,
  pdf_width      = 4 * 3.8,
  pdf_height     = 2 * 3
)

res$thresholds   # per-batch thresholds (Supplementary Table 1)
```

```{r qc-apply}
out <- apply_qc_thresholds(
  seurat_obj    = basak,
  thresholds_df = res$thresholds,
  sample_col    = "qc_batch",
  lowqc_col     = "CellType_AllCells",
  lowqc_groups  = "low_UMI"
)

basak_qc <- out$obj
out$summary           # total=2347, kept=2071, removed=276
out$removed_breakdown

round(ncol(basak_qc) / ncol(basak) * 100, 2)  # 88.24% passed QC
rm(basak)
```


# 5. Re-integration on QC-passed nuclei

```{r reintegrate}
basak_qc[["RNA"]] <- split(basak_qc[["RNA"]], f = basak_qc$Section)

basak_qc <- NormalizeData(basak_qc, verbose = FALSE)
basak_qc <- FindVariableFeatures(basak_qc, verbose = FALSE)
basak_qc <- ScaleData(basak_qc, verbose = FALSE)
basak_qc <- RunPCA(basak_qc, verbose = FALSE)
basak_qc <- IntegrateLayers(basak_qc, method = RPCAIntegration,
                             verbose = FALSE, k.weight = 100)
basak_qc <- RunUMAP(basak_qc, dims = 1:40, reduction = "integrated.dr",
                    verbose = FALSE)
basak_qc <- FindNeighbors(basak_qc, reduction = "integrated.dr",
                           dims = 1:40, k.param = 20, verbose = FALSE)
basak_qc <- FindClusters(basak_qc, resolution = seq(0.1, 1.5, by = 0.1),
                          verbose = FALSE)
basak_qc[["RNA"]] <- JoinLayers(basak_qc[["RNA"]])
```


# 6. Final annotation (QC-passed nuclei)

Annotation at res 0.7 based on canonical marker expression (same marker sets
as the preliminary annotation above; cluster identities verified by dot plot).

```{r final-annotate}
Idents(basak_qc) <- basak_qc$RNA_snn_res.0.7

basak_qc <- RenameIdents(basak_qc,
  "3"  = "Neurons_NE",
  "4"  = "Neurons_NE",
  "11" = "Neurons",
  "7"  = "Neurons",
  "1"  = "Neurons",
  "2"  = "Neurons",
  "5"  = "Neurons",
  "0"  = "Oligo",
  "8"  = "Oligo",
  "9"  = "OPCs",
  "6"  = "Astroependymal",
  "10" = "Microglia"
)

basak_qc$CellType <- Idents(basak_qc)

table(basak_qc$CellType, useNA = "ifany")
# Neurons_NE=463, Neurons=897, Oligo=493, OPCs=56, Astroependymal=131, Microglia=31
```


# 7. Save

```{r save}
basak_qc@meta.data <- basak_qc@meta.data %>%
  dplyr::select(orig.ident, nCount_RNA, log2_nCount_RNA, nFeature_RNA, percent.mt, Section, SectionLevel, qc_batch,
                RNA_snn_res.0.7, CellType, CellType_AllCells)

saveRDS(basak_qc, file = "output/01_basak/basak_snRNA_seurat.rds")
```
