---
title: "02 — snRNA-seq Meta-integration: Three-dataset RPCA, Neuronal Subtypes, and DE"
---

This script integrates the Basak snRNA-seq data with the published Weber et al.
(2024) and Siletti et al. (2023) LC/pons datasets, annotates the integrated
object, performs neuron-level subtype analysis, and runs differential expression
between the two dominant DBH+ NE neuron subtypes.

Inputs:
  - `output/01_basak/basak_snRNA_seurat.rds`
  - `data/external/Weber/weber_snRNA_seurat.rds`
  - `data/external/Siletti/Pons_PnRF_PB_seurat.rds`

Outputs (Supplementary Table 2 documents nuclei counts per dataset):
  - `output/02_meta/meta_snRNA_allcells_seurat.rds`    (46,417 nuclei)
  - `output/02_meta/meta_snRNA_neurons_seurat.rds`     (33,375 neuronal nuclei)
  - `output/02_meta/DE/markers_MAST.csv`

```{r setup}
library(Seurat)
library(SeuratWrappers)
library(dplyr)
library(ggplot2)
library(ggrepel)

dir.create("output/02_meta/DE", recursive = TRUE, showWarnings = FALSE)

options(future.globals.maxSize = 40 * 1024^3)  # 40 GB for large integrations
```


# 1. Prepare datasets for integration

## 1.1 Basak

```{r basak-prep}
basak <- readRDS("output/01_basak/basak_snRNA_seurat.rds")  # 2,071 nuclei

basak <- RenameCells(basak, add.cell.id = "basak")

basak$Dataset <- "Basak"
basak$Sample  <- basak$Section
basak$Donor   <- "Donor_1"
basak$orig.ident <- basak$Section

basak@meta.data <- basak@meta.data %>%
  dplyr::select(Dataset, orig.ident, Donor, Sample,
                nCount_RNA, nFeature_RNA, percent.mt,
                SectionLevel, CellType, CellType_AllCells)
```

## 1.2 Weber et al. 2024

The Weber dataset is available from the original publication.
See `data/README.md` for the expected file path.

```{r weber-prep}
weber <- readRDS("data/external/Weber/weber_snRNA_seurat.rds")  # 20,191 nuclei

weber <- DietSeurat(weber, assay = "RNA")
weber@assays$RNA$scale.data <- NULL
weber@assays$RNA$data       <- NULL

weber$Dataset    <- "Weber"
weber$Donor      <- weber$orig.ident
weber$nCount_RNA   <- weber$nCount_originalexp
weber$nFeature_RNA <- weber$nFeature_originalexp
weber[["percent.mt"]] <- PercentageFeatureSet(weber, pattern = "^MT-")

weber@meta.data <- weber@meta.data %>%
  dplyr::select(Dataset, orig.ident, Donor, Sample,
                nCount_RNA, nFeature_RNA, percent.mt,
                label_merged, supervised_NE, unsupervised_NE)
```

## 1.3 Siletti et al. 2023

Only the parabrachial / pontine reticular formation section (PnRF_PB) containing
DBH+ NE neurons is used. See `data/README.md` for the expected file path.

```{r siletti-prep}
siletti <- readRDS("data/external/Siletti/Pons_PnRF_PB_seurat.rds")  # 24,155 nuclei

siletti$Dataset <- "Siletti"
siletti$Sample  <- as.character(siletti$sample_id)
siletti$Donor   <- as.character(siletti$donor_id)
siletti[["percent.mt"]] <- PercentageFeatureSet(siletti, pattern = "^MT-")

siletti@meta.data <- siletti@meta.data %>%
  dplyr::select(Dataset, orig.ident, Donor, Sample,
                nCount_RNA, nFeature_RNA, percent.mt,
                supercluster_term, cluster_id, subcluster_id,
                dissection, tissue, sex, development_stage)
```


# 2. Three-dataset RPCA integration (all cell types)

Four integration methods were evaluated (RPCA, CCA, Harmony, FastMNN);
RPCA produced the best mixing of datasets while preserving biological structure
and was used for all analyses.

```{r integrate-allcells}
seurat_meta <- merge(basak, y = c(weber, siletti))  # 46,417 nuclei
rm(basak, weber, siletti)

seurat_meta[["RNA"]] <- JoinLayers(seurat_meta[["RNA"]])

seurat_meta <- DietSeurat(seurat_meta, assays = "RNA")
seurat_meta@assays$RNA$scale.data <- NULL

seurat_meta[["RNA"]] <- split(seurat_meta[["RNA"]],
                               f = seurat_meta$Sample)

seurat_meta <- NormalizeData(seurat_meta, verbose = FALSE)
seurat_meta <- FindVariableFeatures(seurat_meta, verbose = FALSE)
seurat_meta <- ScaleData(seurat_meta, verbose = FALSE)
seurat_meta <- RunPCA(seurat_meta, verbose = FALSE)
seurat_meta <- IntegrateLayers(seurat_meta, method = RPCAIntegration,
                                new.reduction = "integrated.rpca",
                                verbose = FALSE, k.weight = 100)
seurat_meta <- FindNeighbors(seurat_meta, reduction = "integrated.rpca",
                              dims = 1:30, k.param = 20, verbose = FALSE)
seurat_meta <- FindClusters(seurat_meta,
                             resolution = seq(0.1, 2.0, by = 0.1),
                             verbose = FALSE)
seurat_meta <- RunUMAP(seurat_meta, reduction = "integrated.rpca",
                        dims = 1:30, reduction.name = "umap.rpca",
                        verbose = FALSE)
seurat_meta[["RNA"]] <- JoinLayers(seurat_meta[["RNA"]])
```

## 2.1 Annotate broad cell types

Annotation at res 0.6 based on canonical marker expression.
NE neurons: DBH, TH, SLC6A2; pan-neuronal: STMN2, SYT1;
oligodendrocytes: MAG, MBP, MOG, PLP1; OPCs: PDGFRA, EGFR;
astrocytes: GFAP, GJA1, AQP4; ependymal: FOXJ1, CCDC153, CAPS;
microglia: C1QB, AIF1; vascular: ACTA2, TAGLN, COL1A1.

```{r annotate-allcells}
Idents(seurat_meta) <- seurat_meta$RNA_snn_res.0.6

seurat_meta <- RenameIdents(seurat_meta,
  "21" = "Neurons_NE",
  "24" = "Neurons",
  "0"  = "Neurons",  "1"  = "Neurons",  "2"  = "Neurons",
  "4"  = "Neurons",  "7"  = "Neurons",  "8"  = "Neurons",
  "10" = "Neurons",  "11" = "Neurons",  "12" = "Neurons",
  "14" = "Neurons",  "15" = "Neurons",  "16" = "Neurons",
  "18" = "Neurons",  "19" = "Neurons",  "20" = "Neurons",
  "23" = "Neurons",  "26" = "Neurons",  "27" = "Neurons",
  "28" = "Neurons",  "29" = "Neurons",
  "3"  = "Oligo",    "5"  = "Oligo",    "9"  = "Oligo",
  "13" = "OPCs",
  "6"  = "Astro",
  "17" = "Microglia",
  "25" = "Ependymo",
  "22" = "Vascular"
)

seurat_meta$CellType_Integrated <- Idents(seurat_meta)

table(seurat_meta$CellType_Integrated, seurat_meta$Dataset, useNA = "ifany")
# Expected (Supplementary Table 2):
#   Neurons_NE : Basak=346, Weber=306, Siletti=113
#   Neurons    : Basak=1012, Weber=16201, Siletti=15397

seurat_meta@assays$RNA$scale.data.1 <- NULL
seurat_meta@assays$RNA$scale.data <- NULL

seurat_meta@meta.data <- seurat_meta@meta.data %>%
  dplyr::select(
    # Core identity
    Dataset, orig.ident, Donor, Sample,
    # Final annotation
    RNA_snn_res.0.6, CellType_Integrated,
    # QC metrics
    nCount_RNA, nFeature_RNA, percent.mt,
    # Basak-specific
    SectionLevel,
    # Weber-specific (useful for cross-dataset comparison)
    label_merged, supervised_NE, unsupervised_NE,
    # Siletti-specific
    supercluster_term, dissection
    )

saveRDS(seurat_meta, file = "output/02_meta/meta_snRNA_allcells_seurat.rds")
rm(seurat_meta)
```


# 3. Neuron-only integration and subtype annotation

```{r neurons-subset}
seurat_meta <- readRDS("output/02_meta/meta_snRNA_allcells_seurat.rds")

neurons <- subset(seurat_meta,
                  idents = c("Neurons_NE", "Neurons"))  # 33,375 nuclei
rm(seurat_meta)

neurons <- DietSeurat(neurons, assays = "RNA")
```

```{r neurons-integrate}
neurons[["RNA"]] <- split(neurons[["RNA"]], f = neurons$Sample)

neurons <- NormalizeData(neurons, verbose = FALSE)
neurons <- FindVariableFeatures(neurons, verbose = FALSE)
neurons <- ScaleData(neurons, verbose = FALSE)
neurons <- RunPCA(neurons, verbose = FALSE)
neurons <- IntegrateLayers(neurons, method = RPCAIntegration,
                            new.reduction = "integrated.rpca",
                            verbose = FALSE, k.weight = 60)
neurons <- FindNeighbors(neurons, reduction = "integrated.rpca",
                          dims = 1:30, k.param = 10, verbose = FALSE)
neurons <- FindClusters(neurons,
                         resolution = seq(0.1, 1.5, by = 0.1),
                         verbose = FALSE)
neurons <- RunUMAP(neurons, reduction = "integrated.rpca",
                    dims = 1:30, reduction.name = "umap.rpca",
                    verbose = FALSE)
neurons[["RNA"]] <- JoinLayers(neurons[["RNA"]])
```

## 3.1 Identify cluster markers using curated GPCR/ligand gene list

Cluster labels are derived by selecting the top-2 discriminative genes from the
IUPHAR curated list per cluster, concatenated into a gene-pair name.

```{r neurons-markers}
Idents(neurons) <- neurons$RNA_snn_res.1.5
length(levels(neurons))  # 68 transcriptionally distinct clusters

# Load IUPHAR GPCR/ligand gene list
gpcr_df    <- read.csv("resources/IUPHAR/GtP_to_HGNC_mapping.csv", skip = 1)
gene_vector <- unique(gpcr_df$HGNC.Symbol)

# Cluster markers restricted to curated gene list
marker_genes <- FindAllMarkers(
  object   = neurons,
  features = intersect(gene_vector, rownames(neurons)),
  verbose  = FALSE
)
write.csv(marker_genes,
          file = "output/02_meta/DEGs_neurons_res1.5_curated.csv")

# Select top 2 discriminative genes per cluster (high FC, low background)
markers_filtered <- marker_genes %>%
  dplyr::filter(avg_log2FC > 0.5 & pct.1 > 0.1 &
                pct.2 < 0.05 & p_val_adj < 0.05) %>%
  group_by(cluster) %>%
  slice_head(n = 2) %>%
  ungroup() %>%
  mutate(cluster = factor(cluster, levels = levels(neurons))) %>%
  arrange(cluster)

# Relax criteria for clusters with no marker meeting strict filters
missing_clusters <- setdiff(levels(neurons),
                             unique(markers_filtered$cluster))

if (length(missing_clusters) > 0) {
  missing_markers <- marker_genes %>%
    dplyr::filter(cluster %in% missing_clusters,
                  avg_log2FC > 0.5 & pct.1 > 0.05 &
                  pct.2 < 0.2 & p_val_adj < 0.05) %>%
    group_by(cluster) %>%
    slice_head(n = 2) %>%
    ungroup() %>%
    mutate(cluster = factor(cluster, levels = levels(neurons))) %>%
    arrange(cluster)
}

marker_df <- rbind(markers_filtered, missing_markers) %>%
  mutate(cluster = factor(cluster, levels = levels(neurons)))
```

## 3.2 Build cluster annotation labels

```{r neurons-annotate}
cluster_annotations <- marker_df %>%
  group_by(cluster) %>%
  summarise(Annotation = paste(gene, collapse = "_"), .groups = "drop") %>%
  mutate(row_id = row_number()) %>%
  arrange(Annotation, row_id) %>%
  group_by(Annotation) %>%
  mutate(
    Annotation_unique = ifelse(
      row_number() == 1,
      Annotation,
      paste0(Annotation, "_", row_number())
    )
  ) %>%
  ungroup() %>%
  arrange(row_id) %>%
  select(-row_id)

write.csv(cluster_annotations,
          file = "output/02_meta/cluster_annotations_neurons.csv")

# Transfer annotations to metadata
seurat_meta_df <- as.data.frame(neurons@meta.data) %>%
  dplyr::select(RNA_snn_res.1.5)

meta_df <- seurat_meta_df %>%
  tibble::rownames_to_column("row_id") %>%
  mutate(Cluster_Join = as.character(RNA_snn_res.1.5)) %>%
  left_join(cluster_annotations, by = c("Cluster_Join" = "cluster")) %>%
  rename(CellType = Annotation_unique) %>%
  select(-Cluster_Join) %>%
  tibble::column_to_rownames("row_id") %>%
  dplyr::select(CellType)

neurons <- AddMetaData(neurons, metadata = meta_df,
                        col.name = "Neuron_Subtype")

# Prioritise DBH subtypes at the top of the level ordering
dbh_clusters  <- c("SLC6A2_DBH", "DBH_SLC6A2", "SLC6A2_MET")
cluster_levels <- c(dbh_clusters,
                    setdiff(sort(unique(neurons$Neuron_Subtype)),
                            dbh_clusters))
neurons$Neuron_Subtype <- factor(neurons$Neuron_Subtype,
                                   levels = cluster_levels)
Idents(neurons) <- neurons$Neuron_Subtype

table(neurons$Neuron_Subtype[neurons$Neuron_Subtype %in% dbh_clusters],
      neurons$Dataset[neurons$Neuron_Subtype %in% dbh_clusters],
      useNA = "ifany")
# SLC6A2/DBH: Basak=256, Weber=285, Siletti=65
# DBH/SLC6A2: Basak=112, Weber=177, Siletti=5
# SLC6A2/MET: Basak=3,   Weber=43,  Siletti=43
```

```{r neurons-save}
neurons@meta.data <- neurons@meta.data %>%
  dplyr::select(
    # Core identity
    Dataset, orig.ident, Donor, Sample,
    # Annotations
    Neuron_Subtype, CellType_Integrated,
    # QC metrics
    nCount_RNA, nFeature_RNA, percent.mt,
    # Basak-specific
    SectionLevel,
    # Weber-specific (useful for cross-dataset comparison)
    label_merged, supervised_NE, unsupervised_NE,
    # Siletti-specific
    supercluster_term, dissection
    )

saveRDS(neurons, file = "output/02_meta/meta_snRNA_neurons_seurat.rds")
rm(neurons)
```


# 4. Differential expression: SLC6A2_DBH vs DBH_SLC6A2 (MAST)

MAST single-cell DE comparing the two most abundant DBH+ NE neuron subtypes,
adjusting for donor effect, library size, and mitochondrial read fraction.

## 4.1 Data preparation

```{r de-prep}
library(MAST)

neurons <- readRDS("output/02_meta/meta_snRNA_neurons_seurat.rds")
Idents(neurons) <- neurons$Neuron_Subtype

dbh_subset <- subset(neurons,
                     idents = c("SLC6A2_DBH", "DBH_SLC6A2"))  # 900 cells
rm(neurons)

# Remove donors with < 20 cells in either subtype (statistical stability)
dbh_subset$Donor <- factor(dbh_subset$Donor)
dbh_subset$Neuron_Subtype <- droplevels(dbh_subset$Neuron_Subtype)

meta_df <- dbh_subset@meta.data %>%
  tibble::rownames_to_column("cell_id") %>%
  group_by(Neuron_Subtype, Donor) %>%
  mutate(n_cells = n()) %>%
  ungroup()

cells_to_keep <- meta_df %>%
  filter(n_cells >= 20) %>%
  pull(cell_id)

dbh_filtered <- subset(dbh_subset, cells = cells_to_keep)  # 890 cells
rm(dbh_subset)

table(dbh_filtered$Neuron_Subtype, dbh_filtered$Donor, useNA = "ifany")
# SLC6A2_DBH: Br2701=157, Br8079=124, Donor_1=256, H18.30.002=41, H19.30.001=24
# DBH_SLC6A2: Br2701=80,  Br8079=96,  Donor_1=112
```

## 4.2 MAST

```{r mast}
Idents(dbh_filtered) <- dbh_filtered$Neuron_Subtype
log2FC_limit <- 2

markers_mast <- FindMarkers(
  object          = dbh_filtered,
  ident.1         = "SLC6A2_DBH",
  ident.2         = "DBH_SLC6A2",
  test.use        = "MAST",
  logfc.threshold = 0.1,
  min.pct         = 0.05,
  only.pos        = FALSE,
  latent.vars     = c("nCount_RNA", "percent.mt", "Donor")
)

write.csv(markers_mast,
          file = "output/02_meta/DE/markers_MAST.csv")

# Construct volcano plot dataframe (used in figures code)
markers_filtered <- markers_mast %>%
  as.data.frame() %>%
  tibble::rownames_to_column("gene") %>%
  mutate(
    neg_log10_p = -log10(p_val_adj),
    sig = case_when(
      p_val_adj < 0.05 & avg_log2FC >  log2FC_limit ~ "Up_in_SLC6A2_DBH",
      p_val_adj < 0.05 & avg_log2FC < -log2FC_limit ~ "Up_in_DBH_SLC6A2",
      TRUE                                           ~ "Not_significant"
    )
  ) %>%
  dplyr::filter(sig %in% c("Up_in_SLC6A2_DBH", "Up_in_DBH_SLC6A2"))

write.csv(markers_filtered,
          file = "output/02_meta/DE/markers_MAST_filtered.csv")

table(markers_filtered$sig)  # Up_in_DBH_SLC6A2 = 20, Up_in_SLC6A2_DBH = 5
rm(dbh_filtered)
```

> HTR2C, SNTG1 : Up_in_DBH_SLC6A2


# 5. FindAllMarkers: DBH subtype markers (used in Fig 1f heatmap)

Seurat FindAllMarkers identifies genes with specificity for each DBH subtype
(including SLC6A2_MET). Genes from this analysis supplement the literature-
curated gene set displayed in the heatmap of Fig 1f.

```{r findallmarkers}
mito_ribo_genes <- readRDS("resources/gene_lists/mito_ribo_genes.rds")

neurons <- readRDS("output/02_meta/meta_snRNA_neurons_seurat.rds")
Idents(neurons) <- neurons$Neuron_Subtype

dbh_subset <- subset(neurons,
  idents = c("SLC6A2_DBH", "DBH_SLC6A2", "SLC6A2_MET"))  # 989 cells
rm(neurons)

# Remove uninformative gene classes
gene_vector <- setdiff(rownames(dbh_subset), mito_ribo_genes)
gene_vector <- setdiff(gene_vector,
                       grep("^ENSG", gene_vector, value = TRUE))
gene_vector <- setdiff(gene_vector,
                       grep("^LINC", gene_vector, value = TRUE))

all_markers <- FindAllMarkers(
  object          = dbh_subset,
  features        = gene_vector,
  logfc.threshold = 0.5,
  min.pct         = 0.05,
  verbose         = FALSE
)
write.csv(all_markers,
          file = "output/02_meta/DE/markers_FindAllMarkers_DBH.csv")

# Filtered marker set used in Fig 1f
markers_filtered <- all_markers %>%
  dplyr::filter(avg_log2FC > 0.5 & pct.1 > 0.2 &
                pct.2 < 0.2 & p_val_adj < 0.05) %>%
  mutate(cluster = factor(cluster,
                           levels = c("SLC6A2_DBH", "DBH_SLC6A2", "SLC6A2_MET"))) %>%
  arrange(cluster)

write.csv(markers_filtered,
          file = "output/02_meta/DE/markers_FindAllMarkers_DBH_filtered.csv")

rm(dbh_subset, all_markers, markers_filtered)
```
> CHRNA3 : Up in SLC6A2_DBH
> LEPR, BDNF, TMC3, INVS : Up in SLC6A2_MET
