---
title: "03 — Basak Visium: Section Loading, QC, and Cropping"
---

This script loads the four Basak Visium sections, computes QC metrics, crops
low-UMI peripheral regions, and saves a merged list of cropped Seurat objects
for downstream integration.

Input:  `data/Visium/<section>/`
Output: `output/03_visium_basak/visium_basak_list.rds` (12,495 spots total)

```{r setup}
library(Seurat)
library(ggplot2)

dir.create("output/03_visium_basak", recursive = TRUE, showWarnings = FALSE)
```


# 1. Load and QC each section

For each section, QC metrics (nCount_RNA, nFeature_RNA, percent.mt) are
visualised and spots in low-UMI peripheral regions are removed by coordinate-
based cropping. Cropping boundaries were determined visually by comparing
`nCount_RNA` spatial plots against tissue position images.

## Upper Rostral

```{r upper-rostral}
L1_1 <- Load10X_Spatial(
  data.dir     = "data/Visium/Upper_Rostral",
  filename     = "filtered_feature_bc_matrix.h5",
  assay        = "RNA",
  slice        = "Up_Rostral",
  filter.matrix = TRUE
)
L1_1[["percent.mt"]] <- PercentageFeatureSet(L1_1, pattern = "^MT-")
L1_1$Section      <- "Upper_Rostral"
L1_1$SectionLevel <- "Rostral"

# Crop low-UMI peripheral region
coords <- L1_1@images$Up_Rostral@coordinates
keep   <- rownames(coords[coords$imagerow > -3300 &
                            coords$imagecol < 33000, ])
L1_1_cropped <- subset(L1_1, cells = keep)
rm(L1_1)
```

## Lower Rostral

```{r lower-rostral}
L1_2 <- Load10X_Spatial(
  data.dir     = "data/Visium/Lower_Rostral",
  filename     = "filtered_feature_bc_matrix.h5",
  assay        = "RNA",
  slice        = "Low_Rostral",
  filter.matrix = TRUE
)
L1_2[["percent.mt"]] <- PercentageFeatureSet(L1_2, pattern = "^MT-")
L1_2$percent.mt[is.na(L1_2$percent.mt)] <- 0
L1_2$Section      <- "Lower_Rostral"
L1_2$SectionLevel <- "Rostral"

coords <- L1_2@images$Low_Rostral@coordinates
keep   <- rownames(coords[coords$imagerow >  -2000 &
                            coords$imagerow < 33000 &
                            coords$imagecol >   1200 &
                            coords$imagecol < 35000, ])
L1_2_cropped <- subset(L1_2, cells = keep)
rm(L1_2)
```

## Upper Caudal

```{r upper-caudal}
L2_3 <- Load10X_Spatial(
  data.dir     = "data/Visium/Upper_Caudal",
  filename     = "filtered_feature_bc_matrix.h5",
  assay        = "RNA",
  slice        = "Up_Caudal",
  filter.matrix = TRUE
)
L2_3[["percent.mt"]] <- PercentageFeatureSet(L2_3, pattern = "^MT-")
L2_3$Section      <- "Upper_Caudal"
L2_3$SectionLevel <- "Caudal"

coords <- L2_3@images$Up_Caudal@coordinates
keep   <- rownames(coords[coords$imagecol < 36500, ])
L2_3_cropped <- subset(L2_3, cells = keep)
rm(L2_3)
```

## Lower Caudal

```{r lower-caudal}
L2_4 <- Load10X_Spatial(
  data.dir     = "data/Visium/Lower_Caudal",
  filename     = "filtered_feature_bc_matrix.h5",
  assay        = "RNA",
  slice        = "Low_Caudal",
  filter.matrix = TRUE
)
L2_4[["percent.mt"]] <- PercentageFeatureSet(L2_4, pattern = "^MT-")
L2_4$percent.mt[is.na(L2_4$percent.mt)] <- 0
L2_4$Section      <- "Lower_Caudal"
L2_4$SectionLevel <- "Caudal"

coords <- L2_4@images$Low_Caudal@coordinates
keep   <- rownames(coords[coords$imagerow > -500, ])
L2_4_cropped <- subset(L2_4, cells = keep)
rm(L2_4)
```


# 2. Spot counts after cropping

```{r spot-counts}
sapply(list(L1_1_cropped, L1_2_cropped, L2_3_cropped, L2_4_cropped), ncol)
# Upper_Rostral  Lower_Rostral  Upper_Caudal  Lower_Caudal
# (totals sum to 12,495 — Supplementary Table 3)
```


# 3. Save

```{r save}
visium_list <- list(
  L1_1 = L1_1_cropped,
  L1_2 = L1_2_cropped,
  L2_3 = L2_3_cropped,
  L2_4 = L2_4_cropped
)

saveRDS(visium_list, file = "output/03_visium_basak/visium_basak_list.rds")
rm(L1_1_cropped, L1_2_cropped, L2_3_cropped, L2_4_cropped, visium_list)
```
