---
title: "04 — Visium Integration: Basak + Weber, Annotation"
---

This script integrates the four Basak Visium sections with the published Weber
et al. (2024) LC Visium data, annotates the resulting spatial clusters, and
saves the final annotated object used for all Visium figures.

Inputs:
  - `output/03_visium_basak/visium_basak_list.rds`
  - `data/external/Weber/weber_visium_seurat.rds`

Output: `output/04_visium_all/meta_visium_seurat.rds` (32,875 spots)

```{r setup}
library(Seurat)
library(dplyr)
library(ggplot2)

dir.create("output/04_visium_all", recursive = TRUE, showWarnings = FALSE)

options(future.globals.maxSize = 40 * 1024^3)
```


# 1. Prepare Weber Visium

The Weber Visium dataset is available from the original publication.
See `data/README.md` for the expected file path.

```{r weber-visium-prep}
weber_visium <- readRDS(
  "data/external/Weber/weber_visium_seurat.rds"
)  # 20,380 spots

weber_visium <- DietSeurat(weber_visium, assays = "RNA")
weber_visium@assays$RNA$scale.data <- NULL
weber_visium@assays$RNA$data       <- NULL

weber_visium$dataset <- "Weber"
weber_visium@meta.data <- as.data.frame(weber_visium@meta.data) %>%
  dplyr::rename(percent.mt = subsets_mito_percent) %>%
  dplyr::select(
    orig.ident, nCount_RNA, nFeature_RNA, percent.mt,
    dataset, sample_id, sample_part_id, donor_id,
    x_coord, y_coord, round_id, part_id,
    array_row, array_col, annot_region, annot_spot, sizeFactor
  )
```


# 2. Prepare Basak Visium

```{r basak-visium-prep}
visium_list <- readRDS("output/03_visium_basak/visium_basak_list.rds")

basak_visium <- merge(
  x           = visium_list[[1]],
  y           = visium_list[2:length(visium_list)],
  add.cell.ids = names(visium_list)
)  # 12,495 spots
rm(visium_list)

# Extract tissue coordinates from all four image slots and store in metadata
# (required to match Weber's x_coord/y_coord format for ggplot-based spatial plots)
image_names <- names(basak_visium@images)

coord_list <- lapply(image_names, function(img) {
  coords <- GetTissueCoordinates(basak_visium[[img]])
  coords$sample_part_id <- img
  coords
})
coords_all <- do.call(rbind, coord_list)
coords_all <- coords_all[rownames(basak_visium@meta.data), , drop = FALSE]

basak_visium@meta.data$x_coord       <- coords_all$imagecol
basak_visium@meta.data$y_coord       <- coords_all$imagerow
basak_visium@meta.data$sample_part_id <- coords_all$sample_part_id

basak_visium <- DietSeurat(basak_visium, assays = "RNA")

basak_visium$sample_id <- basak_visium$Section
basak_visium$dataset   <- "Basak"
basak_visium$donor_id  <- "donor_1"

basak_visium@meta.data <- as.data.frame(basak_visium@meta.data) %>%
  dplyr::select(
    orig.ident, nCount_RNA, nFeature_RNA, percent.mt,
    dataset, sample_id, sample_part_id, donor_id,
    x_coord, y_coord, Section, SectionLevel
  )
```


# 3. Merge and integrate (SCTransform + RPCA)

```{r integrate}
visium_all <- merge(weber_visium, basak_visium)  # 32,875 spots
rm(weber_visium, basak_visium)

visium_all[["RNA"]] <- JoinLayers(visium_all[["RNA"]])
visium_all[["RNA"]] <- split(visium_all[["RNA"]],
                              f = visium_all$sample_part_id)

visium_all <- SCTransform(visium_all, assay = "RNA", verbose = FALSE)
visium_all <- RunPCA(visium_all, verbose = FALSE)

visium_all <- IntegrateLayers(
  visium_all,
  method               = RPCAIntegration,
  normalization.method = "SCT",
  new.reduction        = "integrated.rpca",
  k.weight             = 60,
  verbose              = FALSE
)

visium_all <- FindNeighbors(visium_all, reduction = "integrated.rpca",
                             dims = 1:30, k.param = 10, verbose = FALSE)
visium_all <- FindClusters(visium_all,
                            resolution = seq(0.1, 1.5, by = 0.1),
                            verbose = FALSE)
visium_all <- RunUMAP(visium_all, reduction = "integrated.rpca",
                       dims = 1:30, reduction.name = "umap.rpca",
                       verbose = FALSE)
visium_all <- PrepSCTFindMarkers(visium_all, assay = "SCT", verbose = FALSE)
```


# 4. Identify cluster markers for annotation

Two complementary marker discovery approaches were used:
(a) IUPHAR-curated GPCR/ligand list,
    providing biologically interpretable cell-type identifiers;
(b) genome-wide FindAllMarkers excluding MT/ribosomal genes.
Together these are used to annotate the 17 clusters at res 0.3.

```{r cluster-markers}
# (a) Curated gene list
gpcr_df  <- read.csv("resources/IUPHAR/GtP_to_HGNC_mapping.csv", skip = 1)
gpcr_vector <- sort(unique(gpcr_df$HGNC.Symbol))

Idents(visium_all) <- visium_all$SCT_snn_res.0.3
levels(visium_all) <- sort(as.numeric(levels(visium_all)))

curated_markers <- FindAllMarkers(
  object   = visium_all,
  features = intersect(gpcr_vector, rownames(visium_all)),
  verbose  = FALSE
)
write.csv(curated_markers,
          file = "output/04_visium_all/DEGs_gpcrs_res0.3.csv")

# (b) Genome-wide markers
mito_ribo_genes <- readRDS("resources/gene_lists/mito_ribo_genes.rds")
gene_vector_all <- setdiff(rownames(visium_all), mito_ribo_genes)
gene_vector_all <- setdiff(gene_vector_all,
                            grep("^ENSG", gene_vector_all, value = TRUE))

all_markers <- FindAllMarkers(
  object          = visium_all,
  features        = gene_vector_all,
  logfc.threshold = 0.5,
  min.pct         = 0.1,
  verbose         = FALSE
)
write.csv(all_markers,
          file = "output/04_visium_all/DEGs_all_res0.3.csv")
```


# 5. Annotate clusters

Manual cluster-to-cell-type assignments are stored in
`data/VisiumAll_annotations_res0.3.xlsx`. Annotation was guided by expression
of canonical markers: DBH/TH/SLC6A2 (NE neurons), TPH2/SLC6A4 (serotonergic),
TAC1/CRH, ADCYAP1/VGF, PENK, PVALB/SCN4B, GAD1/GAD2 (GABAergic),
SLC17A6/SLC17A7 (glutamatergic), PLP1/MOG (oligodendrocytes),
AQP4 (astrocytes), FOXJ1/CAPS (ependymal), ACTA2/COL1A1 (vascular).

```{r annotate}
visium_meta <- as.data.frame(visium_all@meta.data) %>%
  dplyr::select(SCT_snn_res.0.3)

# Cluster-to-annotation mapping for Visium resolution 0.3
annot_df <- tibble::tribble(
  ~Cluster, ~Annotation,
  "8",  "NE_Neurons",
  "12", "Seroto_Neurons",
  "14", "TAC1_CRH_Neurons",
  "4",  "ADCYAP1_VGF_Neurons",
  "15", "PENK_Neurons",
  "13", "PVALB_SCN4B_Neurons",
  "3",  "GABA_Neurons",
  "11", "Glut_Neurons",
  "10", "Unclassified_Neurons",   # Weber dataset only
  "6",  "Oligodendrocytes",
  "5",  "Oligodendrocytes",
  "1",  "Oligodendrocytes",
  "2",  "Astrocytes",
  "16", "Ependymal",
  "7",  "Vascular",               # Blood / RBCs
  "9",  "Vascular",               # Endothelial cells
  "0",  "Unclassified"
)

meta_df <- visium_meta %>%
  tibble::rownames_to_column("row_id") %>%
  mutate(Cluster_Join = as.character(SCT_snn_res.0.3)) %>%
  left_join(annot_df, by = c("Cluster_Join" = "Cluster")) %>%
  select(-Cluster_Join) %>%
  tibble::column_to_rownames("row_id") %>%
  dplyr::select(Annotation)

visium_all <- AddMetaData(visium_all, metadata = meta_df)

table(is.na(visium_all$Annotation))  # should be all FALSE

visium_all$Annotation <- factor(visium_all$Annotation,
                                 levels = unique(annot_df$Annotation))
Idents(visium_all) <- visium_all$Annotation

table(visium_all$Annotation, visium_all$dataset, useNA = "ifany")
# Spot counts per annotation × dataset (Supplementary Table 3)
```


# 6. Save annotated object

```{r save}
visium_all@meta.data <- visium_all@meta.data %>%
  dplyr::select(
    # Spatial coordinates
    x_coord, y_coord,
    # Dataset identity
    dataset, donor_id, sample_id, sample_part_id,
    # QC metrics
    nCount_RNA, nFeature_RNA, percent.mt,
    nCount_SCT, nFeature_SCT,
    # Clustering and annotation
    SCT_snn_res.0.3, Annotation,
    # Basak-specific
    Section, SectionLevel,
    # Weber-specific
    part_id, round_id, annot_region, annot_spot, array_row, array_col
  )

visium_all@assays$SCT$scale.data <- NULL

saveRDS(visium_all,
        file = "output/04_visium_all/meta_visium_seurat.rds")
```
