---
title: "Main Figures — Figures 1 and 2"
---

This script produces the panels for main Figures 1 and 2.

Inputs (processed Seurat objects from analysis scripts):
  - `output/01_basak/basak_snRNA_seurat.rds`
  - `output/02_meta/meta_snRNA_allcells_seurat.rds`
  - `output/02_meta/meta_snRNA_neurons_seurat.rds`
  - `output/04_visium_all/meta_visium_seurat.rds`

All plots are saved to `output/figures/main/`.

```{r setup}
library(Seurat)
library(dplyr)
library(ggplot2)

dir.create("output/figures/main", recursive = TRUE, showWarnings = FALSE)
```

```{r color-palettes}
# snRNA-seq Basak cell-type colors (Fig 1b)
basak_snRNA_cols <- c(
  "NE Neurons"       = "#E31A1C",
  "Other Neurons"    = "#FB9A99",
  "Oligodendrocytes" = "#A6CEE3",
  "OPCs"             = "#1F78B4",
  "Astroependymal"   = "#CCEBC5",
  "Microglia"        = "#CAB2D6"
)

# snRNA-seq meta-analysis cell-type colors (Fig 1c, 1d, 1g)
meta_snRNA_cols <- c(
  "NE Neurons"       = "#E31A1C",
  "Other Neurons"    = "#FB9A99",
  "Oligodendrocytes" = "#A6CEE3",
  "OPCs"             = "#1F78B4",
  "Astrocytes"       = "#CCEBC5",
  "Ependymal"        = "grey40",
  "Microglia"        = "#CAB2D6",
  "Vascular"         = "#FED9A6"
)

# DBH subtype colors (Fig 1e, 1f)
dbh_cols <- c(
  "SLC6A2/DBH"   = "#E41A1C",
  "DBH/SLC6A2"   = "#377EB8",
  "SLC6A2/MET"   = "#4DAF4A",
  "Other Neurons" = "#D9D9D9"
)

# Visium cell-type colors (Fig 2b)
visium_cols <- c(
  "NE Neurons"           = "#E31A1C",
  "Serotonergic Neurons" = "#FF7F00",
  "TAC1/CRH Neurons"     = "#6A3D9A",
  "ADCYAP1/VGF Neurons"  = "#FDBF6F",
  "PENK Neurons"         = "#E7298A",
  "PVALB/SCN4B Neurons"  = "#B15928",
  "GABAergic Neurons"    = "#33A02C",
  "Glutamatergic Neurons"= "#B2DF8A",
  "Unclassified Neurons" = "#FB9A99",
  "Oligodendrocytes"     = "#A6CEE3",
  "Astrocytes"           = "#CCEBC5",
  "Ependymal"            = "grey40",
  "Vascular"             = "#FED9A6",
  "Unclassified"         = "#D9D9D9"
)
```


# Figure 1b — Basak snRNA-seq UMAP

```{r fig1b}
basak <- readRDS("output/01_basak/basak_snRNA_seurat.rds")

basak <- RenameIdents(basak,
  "Neurons_NE"    = "NE Neurons",
  "Neurons"       = "Other Neurons",
  "Oligo"         = "Oligodendrocytes",
  "OPCs"          = "OPCs",
  "Astroependymal" = "Astroependymal",
  "Microglia"     = "Microglia"
)

p_1b <- DimPlot(
  object = basak, 
  cols = basak_snRNA_cols, 
  pt.size = 1.5,
  label = F
) + 
  theme_classic(base_family = "Times New Roman", base_size = 12) + 
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  ) +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(colour = "black"),
    legend.position = "right",
    legend.text = element_blank(),
    legend.key.size = unit(9, "mm")
  ) + 
  guides(
    colour = guide_legend(override.aes = list(size = 5))
    )

ggsave(
  plot     = p_1b,
  filename = "output/figures/main/Fig1b_UMAP_basak_snRNA.png",
  height   = 3.5, width = 4.5, units = "in", dpi = 1200
)

rm(basak)
```


# Figure 1c — Meta-analysis UMAP (cell types)

```{r fig1c}
meta <- readRDS("output/02_meta/meta_snRNA_allcells_seurat.rds")
Idents(meta) <- meta$CellType_Integrated

# Re-label for display
meta <- RenameIdents(meta,
  "Neurons_NE" = "NE Neurons",
  "Neurons"    = "Other Neurons",
  "Oligo"      = "Oligodendrocytes",
  "OPCs"       = "OPCs",
  "Astro"      = "Astrocytes",
  "Ependymo"   = "Ependymal",
  "Microglia"  = "Microglia",
  "Vascular"   = "Vascular"
)

p_1c <- DimPlot(
  object = meta, 
  cols = meta_snRNA_cols, 
  pt.size = 0.1,
  label = F
) + 
  theme_classic(base_family = "Times New Roman", base_size = 16) + 
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  ) +
  theme(
    axis.title = element_text(),
    axis.text = element_text(colour = "black"),
    legend.position = "right",
    legend.text = element_blank(),
    legend.key.size = unit(10, "mm")
  ) + 
  guides(
    colour = guide_legend(override.aes = list(size = 6))
    )

ggsave(
  plot     = p_1c,
  filename = "output/figures/main/Fig1c_UMAP_meta_snRNA.png",
  height   = 5, width = 6.5, units = "in", dpi = 1200
)
```


# Figure 1d — Dot plot: canonical cell-type markers

```{r fig1d}
levels(meta) <- rev(names(meta_snRNA_cols))

ct_markers <- c(
  # Pan-neuronal
  "STMN2", "SYT1",
  # NE
  "DBH", "TH", "SLC6A2",
  # Oligodendrocytes
  "MOG", "PLP1", "MAG",
  # OPCs
  "PDGFRA", "OLIG1", "OLIG2",
  # Astrocytes
  "GFAP", "AQP4", "AGT",
  # Ependymal
  "FOXJ1", "CCDC153", "CAPS",
  # Microglia
  "P2RY12", "C1QB", "AIF1",
  # Vascular
  "FLT1", "DCN", "HBA2"
)

p_1d <- DotPlot(meta,
                features = ct_markers,
                assay    = "RNA",
                scale    = TRUE) +
  theme_classic(base_family = "Times New Roman", base_size = 12) + 
  labs(
    x = "Gene",
    y = element_blank()
  ) + 
  theme(
    axis.title = element_text(size = 10.5, colour = "black"),
    axis.text.x = element_text(colour = "black", angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(colour = "black", face = "bold"),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.1, "in")
  ) + 
  guides(
    color = guide_colorbar(title = "Scaled\nExpression"),
    size = guide_legend(title = "Percent\nExpressed")
  )

ggsave(
  plot     = p_1d,
  filename = "output/figures/main/Fig1d_DotPlot_ct_markers.png",
  height   = 2.8, width = 6.7, units = "in", dpi = 1200
)

rm(meta)
```


# Figure 1e — Neurons UMAP (DBH subtypes highlighted)

```{r fig1e}
neurons <- readRDS("output/02_meta/meta_snRNA_neurons_seurat.rds")
levels(neurons) <- rev(levels(neurons))

dbh_map <- c(
  "SLC6A2_DBH" = "SLC6A2/DBH",
  "DBH_SLC6A2" = "DBH/SLC6A2",
  "SLC6A2_MET" = "SLC6A2/MET"
)

clust    <- levels(neurons)
new_names <- setNames(
  ifelse(clust %in% names(dbh_map), dbh_map[clust], "Other Neurons"),
  clust
)
neurons <- RenameIdents(neurons, new_names)

p_1e <- DimPlot(neurons,
                reduction = "umap.rpca",
                cols      = dbh_cols,
                pt.size   = 0.1,
                label     = FALSE) +
  theme_classic(base_family = "Times New Roman", base_size = 16) + 
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  ) +
  theme(
    axis.title = element_text(),
    axis.text = element_text(colour = "black"),
    legend.position = "none"
  ) + 
  guides(
    colour = guide_legend(override.aes = list(size = 6))
    )

ggsave(
  plot     = p_1e,
  filename = "output/figures/main/Fig1e_UMAP_neurons_DBH_subtypes.png",
  height   = 5, width = 6, units = "in", dpi = 1200
)
```


# Figure 1f — Heatmap: average expression across DBH subtypes

Gene selection combines NE core markers, literature-curated subtype markers
(from Weber et al. 2023 and cited references), and DE hits from MAST and
FindAllMarkers (see `02_meta_integration.Rmd`).

```{r fig1f}
dbh_subset <- subset(neurons, idents = unname(dbh_map))
levels(dbh_subset) <- unname(dbh_map)
rm(neurons)

# Gene set for Fig 1f (NE core + literature-curated + DE hits)
gene_vector <- unique(c(
  # NE core
  "DBH", "SLC6A2", "TH",
  # Curated subtype markers (one or two subtypes)
  "NPY",                             # SLC6A2/DBH
  "HTR2C", "SNTG1",                 # DBH/SLC6A2
  "LEPR", "BDNF", "TMC3", "INVS",  # SLC6A2/MET
  # Shared across two subtypes (literature-curated)
  "GAL", "CARTPT", "GPX3", "DLK1", "CHRNA3",
  "MET"
))
genes_present <- intersect(gene_vector, rownames(dbh_subset))

# Average expression per cluster (unscaled, for display)
avg_exp <- DotPlot(dbh_subset, features = genes_present, scale = FALSE)$data

avg_exp$features.plot <- factor(avg_exp$features.plot,
                                levels = genes_present)
avg_exp$id <- factor(avg_exp$id,
                     levels = rev(unname(dbh_map)))

p_1f <- ggplot(avg_exp,
               aes(x = features.plot, y = id, fill = avg.exp.scaled)) +
  geom_tile(color = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low      = "white",
    mid      = "grey85",
    high     = "darkblue",
    midpoint = 0,
    name     = "Average\nExpression"
  ) +
  coord_fixed() +
  theme_classic(base_family = "Times New Roman", base_size = 12) +
  labs(x = "Gene", y = element_blank()) +
  theme(
    axis.title   = element_text(size = 16),
    axis.text.x  = element_text(colour = "black", face = "bold",
                                size = 14, angle = 45, vjust = 1, hjust = 1),
    axis.text.y  = element_text(colour = "black", face = "bold", size = 15),
    legend.title = element_text(size = 14),
    legend.text  = element_text(size = 13),
    legend.key.size = unit(0.2, "in"),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )

ggsave(
  plot     = p_1f,
  filename = "output/figures/main/Fig1f_Heatmap_DBH_subtypes.png",
  height   = 3.5, width = 10, units = "in", dpi = 1200,
  bg       = "transparent"
)

rm(dbh_subset, avg_exp)
```


# Figure 1g — Dot plot: melanocortin pathway and AD-associated genes

```{r fig1g}
meta <- readRDS("output/02_meta/meta_snRNA_allcells_seurat.rds")
neurons <- readRDS("output/02_meta/meta_snRNA_neurons_seurat.rds")

# Transfer neuron subtype labels into the all-cells object
neurons_df <- as.data.frame(neurons@meta.data) %>%
  dplyr::select(Neuron_Subtype)
meta <- AddMetaData(meta, metadata = neurons_df)

# Build combined annotation: DBH subtypes + broad cell types
meta$Annot <- ifelse(
  meta$Neuron_Subtype %in% c("SLC6A2_DBH", "SLC6A2_MET", "DBH_SLC6A2"), 
  as.character(meta$Neuron_Subtype), 
  as.character(meta$CellType_Integrated)
  )
meta$Annot <- as.character(meta$Annot)
meta$Annot <- ifelse(meta$Annot %in% c("SLC6A2_DBH"), "SLC6A2/DBH", meta$Annot)
meta$Annot <- ifelse(meta$Annot %in% c("DBH_SLC6A2"), "DBH/SLC6A2", meta$Annot)
meta$Annot <- ifelse(meta$Annot %in% c("SLC6A2_MET"), "SLC6A2/MET", meta$Annot)
meta$Annot <- ifelse(meta$Annot %in% c("Neurons_NE", "Neurons"), "Other Neurons", meta$Annot)
meta$Annot <- ifelse(meta$Annot %in% c("Oligo"), "Oligodendrocytes", meta$Annot)
meta$Annot <- ifelse(meta$Annot %in% c("Astro"), "Astrocytes", meta$Annot)
meta$Annot <- ifelse(meta$Annot %in% c("Ependymo"), "Ependymal", meta$Annot)

ct_order <- c("SLC6A2/DBH", "DBH/SLC6A2", "SLC6A2/MET", "Other Neurons", 
             "Oligodendrocytes", "OPCs", "Astrocytes", "Ependymal", "Microglia", "Vascular")

meta$Annot <- factor(meta$Annot, levels = ct_order)

Idents(meta) <- meta$Annot
levels(meta) <- rev(levels(meta))

melanocortin_ad_genes <- c(
  # Melanocortin pathway
  "LEPR", "MC1R", "MC2R", "MC3R", "MC4R", "MC5R", "MRAP2",
  # AD-associated
  "APP", "PSEN1", "PSEN2", "APOE", "SORL1", "TREM2"
)

p_1g <- DotPlot(meta,
                features = melanocortin_ad_genes,
                idents = setdiff(ct_order, "Other Neurons"),
                assay    = "RNA",
                scale    = TRUE) +
  theme_classic(base_family = "Times New Roman", base_size = 12) + 
  labs(
    x = "Gene",
    y = element_blank()
  ) + 
  theme(
    axis.title = element_text(size = 10.5, colour = "black"),
    axis.text.x = element_text(colour = "black", angle = 45, hjust = 1, vjust = 1, face = "bold"),
    axis.text.y = element_text(colour = "black", face = "bold"),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.1, "in")
  ) + 
  guides(
    color = guide_colorbar(title = "Scaled\nExpression"),
    size = guide_legend(title = "Percent\nExpressed")
  )

ggsave(
  plot     = p_1g,
  filename = "output/figures/main/Fig1g_DotPlot_melanocortin_AD.png",
  height   = 2.8, width = 6.7, units = "in", dpi = 1200
)

rm(meta, neurons)
```


# Figure 2b — Visium: spatial cell-type map (Lower Rostral section)

```{r fig2b}
visium_all <- readRDS("output/04_visium_all/meta_visium_seurat.rds")

visium_all <- RenameIdents(visium_all,
  "NE_Neurons"           = "NE Neurons",
  "Seroto_Neurons"       = "Serotonergic Neurons",
  "TAC1_CRH_Neurons"     = "TAC1/CRH Neurons",
  "ADCYAP1_VGF_Neurons"  = "ADCYAP1/VGF Neurons",
  "PENK_Neurons"         = "PENK Neurons",
  "PVALB_SCN4B_Neurons"  = "PVALB/SCN4B Neurons",
  "GABA_Neurons"         = "GABAergic Neurons",
  "Glut_Neurons"         = "Glutamatergic Neurons",
  "Unclassified_Neurons" = "Unclassified Neurons",
  "Oligodendrocytes"     = "Oligodendrocytes",
  "Astrocytes"           = "Astrocytes",
  "Ependymal"            = "Ependymal",
  "Vascular"             = "Vascular",
  "Unclassified"         = "Unclassified"
)

# Lower Rostral representative section
p_2b <- SpatialDimPlot(
  visium_all,
  images = "Low_Rostral",
  image.alpha = 0,
  pt.size.factor = 1.5,
  cols = visium_cols
) + 
  scale_y_reverse() +
  theme_classic(base_family = "Times New Roman", base_size = 12) +
  theme(
    plot.title = element_text(),
    text = element_text(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    legend.key.size = unit(0.1, "in"), 
    legend.key.height = unit(0.25, "in"),
    legend.title = element_blank(),
    legend.text = element_text(size = 16)
  ) +
  guides(
    fill = guide_legend(override.aes = list(size = 5))
    )

ggsave(
  plot     = p_2b,
  filename = "output/figures/main/Fig2b_SpatialPlot_LowerRostral_celltypes.png",
  height   = 4, width = 6.5, units = "in", dpi = 1200
)
```


# Figure 2c — Visium: spatial gene expression (Lower Rostral section)

```{r fig2c}
DefaultAssay(visium_all) <- "RNA"
visium_all <- NormalizeData(visium_all, verbose = FALSE)

genes_2c <- c("DBH", "MRAP2", "APP", "PSEN1",
              "PSEN2", "APOE", "SORL1", "TREM2")

for (gene in genes_2c) {

   # Pull expression values to calculate dynamic breaks
  expr_vals <- FetchData(visium_all, vars = gene, slot = "data")[,1]
  min_val <- floor(min(expr_vals))
  max_val <- ceiling(max(expr_vals))
  mid_val <- (min_val + max_val) / 2
  
  # Generate plot
  p <- SpatialFeaturePlot(visium_all, images = "Low_Rostral", image.alpha = 0, 
                          pt.size.factor = 1.5, features = gene) + 
    scale_y_reverse() + 
    labs(
      title = gene
    ) +
    theme_void(base_family = "Times New Roman", base_size = 13) +
    theme(
      plot.title = element_text(size = 20, hjust = 0.5, face = "bold.italic"), 
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.line = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "right",
      legend.key.size = unit(0.15, "in"),
      legend.key.width = unit(0.18, "in"),
      legend.title = element_blank(),
      legend.text = element_text(size = 15),
      panel.background = element_rect(fill = "transparent", colour = NA),  
      plot.background = element_rect(fill = "transparent", colour = NA)
      ) +
    # Force the legend to only show 3 specific integer/rounded breaks
    scale_fill_gradientn(
      colors = rev(RColorBrewer::brewer.pal(n = 11, name = "Spectral")),
      breaks = c(min_val, mid_val, max_val),
      labels = c(min_val, mid_val, max_val),
      limits = c(min_val, max_val) # Explicitly set limits to prevent clipping
    )

  ggsave(
    plot     = p,
    filename = paste0("output/figures/main/Fig2c_SpatialExpr_LowerRostral_",
                      gene, ".png"),
    height   = 2.6, width = 3.1, units = "in", dpi = 1200, bg = "transparent"
  )
}

rm(visium_all)
```
